---
title: "AFOLU and Biodiversity"
author: "Marcus Thomson"
date: '2022-02-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r, message=FALSE}
library(tidyverse)
library(raster)
library(sf)
library(maps)
```


## Load data
### Biodiversity
Species richness for all species.
Source: IUCN. (2022). IUCN Redlist database. <https://www.iucnredlist.org> Retrieved: 25/02/2022.
```{r}
biodiv_data <- raster("data/Richness_all.tif")
```

### Country political borders
Shapefiles for country-level political boundaries (borders).
Source: <https://public.opendatasoft.com/explore/dataset/world-administrative-boundaries>
NB: Should update this to use a newer method.
```{r, message=FALSE}
borders_only <- readOGR(dsn = "data/world-administrative-boundaries", layer = "world-administrative-boundaries")
```

## Pathways & historical data
Historical data from the FAO.
Source: FAO.
```{r, show_col_types=FALSE}
pathways <- readr::read_csv("data/food_feed_pathways.csv")
hist_fao <- readr::read_csv("data/food_feed_fao.csv")
hist_fao <- hist_fao %>%
   dplyr::rename(fao_countrycode = area_code)
```

## Map country names to ISO codes
DF to map ISO3 codes to country names.
```{r}
country_code_map <- readr::read_csv("data/country_mapper.csv")
```

Reorganize the historical FAO data to make it more alike the projected pathways data frame.
```{r}
historical_pathways <- country_code_map %>%
  dplyr::right_join(
    hist_fao, 
    by = "fao_countrycode"
  ) %>% 
  dplyr::mutate(
    diet_plan = "Hist-FAO", 
    scenario = "historical", 
    kt_yr_se = NA, 
    target_year = NA
  ) %>% 
  dplyr::select(iso_alpha3, diet_plan, scenario, food_group, year, kt_yr, kt_yr_smoothed, diff1, diff2)
```

## Crop and mask biodiversity raster
Make a list of extant country iso3 codes.
```{r}
country_iso3_codes <- borders_only@data$iso3[!is.na(borders_only@data$iso3)]
country_iso3_codes <- as.character(country_iso3_codes[order(country_iso3_codes)])
country_iso3_df <- data.frame(index = country_iso3_codes %>% dplyr::row_number(), 
           iso3 = country_iso3_codes)
```

Crop and then mask `biodiv_data` by individual country border polygons.
```{r}
borders_only_sub <-
  borders_only[as.character(borders_only@data$iso3) %in% country_iso3_df$iso3,]
```

Loop over iso3 codes to make a list of biodiveristy rasters for each country. (Bound to be a slicker way of doing this.)
```{r}
biodiv_country_rast <- list()
for (country_iso3 in country_iso3_codes[1:10]) {
  biodiv_sub <-
    crop(biodiv_data, extent(borders_only_sub[borders_only_sub$iso3 == country_iso3, ]))
  biodiv_country_rast[[country_iso3]] <-
    mask(biodiv_sub, borders_only_sub[borders_only_sub$iso3 == country_iso3, ])
}
```

Extract some basic data (mean, sd, median, min, max) from each of these.
```{r}
get_raster_summary <- function(x) {
  tmp1 <- raster::values(x) %>%
    na.omit() %>%
    summary() %>%
    as.vector()
  
  tmp2 <- raster::values(x) %>%
    na.omit() %>%
    sd()
  
  return(data.frame(
    names = c(
      "min",
      "first_quart",
      "median",
      "mean",
      "third_quart",
      "max",
      "sd"
    ),
    biodiv = c(tmp1, tmp2)
  ))
}

biodiv_country_moments <-
  lapply(biodiv_country_rast, FUN = get_raster_summary)
```


```{r}
tmp1 <- pathways %>%
  dplyr::filter(
    target_year == 2050 &
      iso_alpha3 == "AFG" &
      food_group == "animal products" &
      scenario == "SSP1" & diet_plan == "CNS-CFP"
  ) %>%
  dplyr::pull(diff1) %>%
  na.omit() 
```


```{r}
tmp2 <- historical_pathways %>%
  dplyr::filter(
      iso_alpha3 == "AFG" &
      food_group == "animal products"
  ) %>%
  dplyr::pull(diff1) %>%
  na.omit() 
```


```{r}

ggplot(
  data = pathways %>%
    dplyr::filter(
      iso_alpha3 == "AFG" &
        food_group == "plant products" &
        diet_plan == "CNS-CFP" & 
        scenario == "SSP3"
    ),
  aes(x = diff2)
) +
  geom_density(aes(fill = as.factor(target_year)), 
               alpha = 0.5) + 
  geom_density(data = historical_pathways %>%
    dplyr::filter(
      iso_alpha3 == "AFG" &
        food_group == "plant products"), 
               mapping = aes(x = diff2), 
               alpha = 0.1)

```



```{r}

```




## Including Plots

You can also embed plots, for example:

```{r, echo=FALSE}

```


